---
title: "Chapter 4 - Data Visualization Primer"
format: html
---
# Load Libraries

To begin we must load some libraries we will be using. If we do not load them, R will not be able to find the functions contained in these libraries. The tidyverse includes ggplot and other tools. We also load the socviz and gapminder libraries.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(gapminder)
library(here)
library(tidyverse)
library(socviz)
```


### 4.2 Grouped data and the group aesthetic

```{r}
#| title: "An incorrect plot"

p <- ggplot(data = gapminder,
            mapping = aes(x = year,
                          y = gdpPercap))
p + geom_line()       
```

```{r}
p <- ggplot(data = gapminder,
            mapping = aes(x = year,
                          y = gdpPercap))
p + geom_line(mapping = 
                    aes(group = country))       
```


### 4.3 Facet to make small multiples 

```{r facet_1}
p <- ggplot(data = gapminder,
            mapping = aes(x = year,
                          y = gdpPercap))

p + geom_line(mapping =  
              aes(group = country)) + 
facet_wrap(~ continent)      
```

```{r}
p + geom_line(color="gray70",
              mapping=aes(group = country)) +
    geom_smooth(size = 1.1,
                method = "loess",
                se = FALSE) +
    scale_y_log10(labels=scales::dollar) +
    facet_wrap(~ continent, ncol = 5) +
    labs(x = "Year",
         y = "log GDP per capita",
         title = "GDP per capita on Five Continents")      
```


```{r}

p <- ggplot(data = gss_sm,
            mapping = aes(x = age, y = childs))
p + geom_point(alpha = 0.2) + geom_smooth() +
    facet_grid(sex ~ race)

```

### 4.4 Geoms can transform data

```{r}

p <- ggplot(data = gss_sm,
            mapping = aes(x = bigregion))
p + geom_bar()

```

```{r}

p <- ggplot(data = gss_sm,
            mapping = aes(x = bigregion))
p + geom_bar(mapping = aes(y = ..prop..))

```

```{r}

p <- ggplot(data = gss_sm,
            mapping = aes(x = bigregion))
p + geom_bar(mapping = aes(y = ..prop.., group = 1)) 


```

### 4.5 Frequency Plots the Slightly Awkward Way

```{r}

table(gss_sm$religion)

```

```{r}


p <- ggplot(data = gss_sm,
            mapping = aes(x = religion, color = religion))
p + geom_bar()

p <- ggplot(data = gss_sm,
            mapping = aes(x = religion, fill = religion))
p + geom_bar() + guides(fill = FALSE)

```

```{r}

p <- ggplot(data = gss_sm,
            mapping = aes(x = bigregion, 
                          fill = religion))
p + geom_bar()

```

```{r}
p <- ggplot(data = gss_sm,
            mapping = aes(x = bigregion, 
                          fill = religion))
p + geom_bar(position = "fill")      
```

```{r}
p <- ggplot(data = gss_sm,
            mapping = aes(x = bigregion, 
                          fill = religion))
p + geom_bar(position = "dodge",
             mapping = aes(y = ..prop..))      
```


```{r}
p <- ggplot(data = gss_sm,
            mapping = aes(x = bigregion, 
                          fill = religion))
p + geom_bar(position = "dodge",
             mapping = aes(y = ..prop.., 
                           group = religion))       
```

### 4.6 Histograms and Density Plots 

```{r}

p <- ggplot(data = midwest,
            mapping = aes(x = area))
p + geom_histogram()

p <- ggplot(data = midwest,
            mapping = aes(x = area))
p + geom_histogram(bins = 10)

```


```{r}

oh_wi <- c("OH", "WI")

p <- ggplot(data = subset(midwest, subset = state %in% oh_wi),
            mapping = aes(x = percollege, fill = state))
p + geom_histogram(alpha = 0.4, bins = 20)

```


```{r}

p <- ggplot(data = midwest,
            mapping = aes(x = area))
p + geom_density()


```

```{r}


p <- ggplot(data = midwest,
            mapping = aes(x = area, fill = state, color = state))
p + geom_density(alpha = 0.3)

```

### 4.7 Avoid transformations when necessary


```{r}

titanic

```

```{r}

p <- ggplot(data = titanic,
            mapping = aes(x = fate, y = percent, fill = sex))
p + geom_bar(position = "dodge", stat = "identity") + theme(legend.position = "top")

```


```{r}

oecd_sum

```

```{r}

p <- ggplot(data = oecd_sum,
            mapping = aes(x = year, y = diff, fill = hi_lo))
p + geom_col() + guides(fill = FALSE) +
  labs(x = NULL, y = "Difference in Years",
       title = "The US Life Expectancy Gap",
       subtitle = "Difference between US and OECD
                   average life expectancies, 1960-2015",
       caption = "Data: OECD. After a chart by Christopher Ingraham,
                  Washington Post, December 27th 2017.")

```
